---
title: "Introduction to Data Science Assignment 03"
author: "Ryan Collins"
date: "2/14/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

Here are the libraries which will be needed:

```{r load-packages, message = FALSE}
library(readxl)
library(tidyverse)
```

I started by figuring out which data set to use. I ended up downloading Census data that focused on assets at the household level to analyzee the wealth by different households and types of assets they held. To do this I needed to read the Excel file from the Census website, which can be [found here](https://www.census.gov/data/tables/2015/demo/wealth/wealth-asset-ownership.html)

Once I had the data, I needed to move forward with subsectioning the data to get focus on several areas. The code I used is below.

```{r }
#Read in the master dataset:
wealth_tables_cy2015 <- read_excel("wealth_tables_cy2015.xlsx")

```

```{r}
#subsection dataset into area of interest:

#Regions in the US:
regions <- filter(wealth_tables_cy2015, Category == "Region")

#Education levels:
edu <- filter(wealth_tables_cy2015, Category == "Highest Level of Educational Attainment in the Household")	

#Household Networth:
hh_networth <- filter(wealth_tables_cy2015, Category == "Household Net Worth")

#If Individual's are assisted by government programs:
gov_program <- filter(wealth_tables_cy2015, Category == "Participation in Government Programs")

#Ethnicity:
ethnicity <- filter(wealth_tables_cy2015, Category == "Race and Hispanic Origin of Householder")

```

With this done I can start my first plot which looks at the median networth of housholds by ethnicity:

```{r make-plot, fig.width= 12, fig.height=8, eval= TRUE,  echo = FALSE}
ggplot(ethnicity, mapping = aes(x = Characteristic, y = Networth)) +
  geom_bar(aes(fill = "red", color = "black"), stat = "identity") + 
             labs(title = "Median Networth by Ethnicity",
                  x = "Ethnicty",
                  y = "Median Networth",
       caption = "Source: Census Bureau") +
  scale_color_manual(values = "#190707", ) +
  scale_fill_manual(values = "#08088A") + 
  theme(
    plot.title = element_text(face = "bold", size = 10),
    axis.line = element_line(color = "black"),
    legend.position = "none")

```
