---
title: "Introduction to Data Science Assignment 03"
author: "Ryan Collins"
date: "2/14/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
```

Here are the libraries which will be needed:

```{r load-packages, message = FALSE}
library(readxl)
library(tidyverse)
```

I started by figuring out which data set to use. I ended up downloading Census data that focused on assets at the household level to analyzee the wealth by different households and types of assets they held. To do this I needed to read the Excel file from the Census website, which can be [found here](https://www.census.gov/data/tables/2015/demo/wealth/wealth-asset-ownership.html)

Once I had the data, I needed to move forward with subsectioning the data to get focus on several areas. The code I used is below.

```{r}
#Read in the master dataset:
wealth_tables_cy2015 <- read_excel("wealth_tables_cy2015.xlsx")
#Also importing an additional dataset here:
fed_hh_balance <- read_csv("z1-visualization-data.csv")
#there is an additional step to rename all the variables which can be seen in the r code but 
#not displayed here.

```

```{r include = FALSE}
fed_hh_balance <- fed_hh_balance %>%
  rename(hh.npo.net.worth = FL152090005.Q,
         hh.npo.net.worth.DPI.percentage = FL152090006.Q,
         hh.npo.total.assets = FL152000005.Q,
         hh.npo.total.assets.dpi.percentage = FL152000006.Q,
         hh.npo.total.liabilities = FL154190005.Q,
         hh.npo.total.liabilities.dpi.percentage = FL154190006.Q,
         hh.npo.nonfinancial.assets = LM152010005.Q,
         hh.npo.household.nonfinancial.assets.dpi.percentage = LM152010006.Q,
         hh.npo.financial.assets = FL154090005.Q,
         hh.npo.financial.assets.dpi.percentage = FL154090006.Q,
         hh.npo.home.mortgages = FL153165105.Q,
         hh.npo.home.mortgages.dpi.percentage = FL153165106.Q,
         hh.npo.consumer.credit = FL153166000.Q,
         hh.npo.consumer.credit.dpi.percentage = FL153166006.Q,
         hh.npo.other.liabilities= FL154199005.Q,
         hh.npo.pther.liabilities.dpi.percentage = FL154199006.Q,
         hh.npo.real.estate = LM155035015.Q,
         hh.npo.Real.estate.dpi.percentage = LM155035016.Q,
         hh.npo.consumer.durables = LM155111005.Q,
         hh.npo.consumer.durables.dpi.percentage = LM155111006.Q,
         npo.nonfinancial.assets = LM162010005.Q,
         npo.nonfinancial.assets.dpi.percentage = LM162010006.Q,
         hh.npo.deposits = FL154000025.Q,
         hh.npo.deposits.dpi.percentage = FL154000026.Q,
         hh.npo.directly.held.corporate.equities = LM153064105.Q,
         hh.npo.directly.held.corporate.equities.dpi.percentage = LM153064106.Q,
         hh.npo.indirectly.held.corporate.eqities = LM153064175.Q,
         hh.npo.indirectly.held.corporate.equities.dpi.percentage = LM153064176.Q,
         hhdbpe = FL594190045.Q,
         hhdbpe.dpi.percentage = FL594190046.Q,
         hh.npo.proprietors.equity.in.noncorporate.business = LM152090205.Q,
         hh.npo.proprietors.equity.in.noncorporate.business.dpi.percentage = LM152090206.Q,
         hh.npo.other.financial_assets = FL153099005.Q,
         hh.npo.other.financial.assets.dpi.percentage = FL153099006.Q,
         hh.npo.net.worth.change.in.unadjusted.level = FC152090005.Q,
         hh.npo.direct.and.indirect.holdings.of.corporate.equity.change.in.unadjusted_level = FC153064475.Q,
         hh.npo.real.estate.change.in.unadjusted.level = FC155035005.Q,
         hh.npo.other.change.in.unadjusted.level = FC152090045.Q,
         disposable.personal.income = FA156012005.Q)
```

```{r}
#subsection both datasets into area of interest:

#Regions in the US:
regions <- filter(wealth_tables_cy2015, Category == "Region")
#Education levels:
edu <- filter(wealth_tables_cy2015, Category == "Highest Level of Educational Attainment in the Household")	
#Household Networth:
hh_networth <- filter(wealth_tables_cy2015, Category == "Household Net Worth")
#If Individual's are assisted by government programs:
gov_program <- filter(wealth_tables_cy2015, Category == "Participation in Government Programs")
#Ethnicity:
ethnicity <- filter(wealth_tables_cy2015, Category == "Race and Hispanic Origin of Householder")
#Household data from 2000-2019:
fed_hh_balance_2000 <- filter(fed_hh_balance, date >= "2000:Q1")

```

With this done I can start my first plot which looks at the median networth of housholds by ethnicity:

```{r plot1, fig.width= 12, fig.height=8, eval = TRUE, echo = FALSE}
ggplot(ethnicity, mapping = aes(x = Characteristic, y = Networth)) +
  geom_bar(aes(fill = "red", color = "black"), stat = "identity") + 
             labs(title = "Median Networth by Ethnicity",
                  x = "Ethnicty",
                  y = "Median Networth",
       caption = "Source: Census Bureau") +
  scale_color_manual(values = "#190707", ) +
  scale_fill_manual(values = "#08088A") + 
  theme(
    plot.title = element_text(face = "bold", size = 10),
    axis.line = element_line(color = "black"),
    legend.position = "none")

```

And a second visualizaiton that looks at the corrleation between median totals in houshold retirement accounts and net worth based upon educational attainment. I've also included a third variable in this plot, illustrating home equity: 


```{r plot2, fig.width= 12, fig.height=8, warning = FALSE,  echo = FALSE}
ggplot(edu, mapping = aes(x = Networth, y = Retirement_accounts_Total)) + 
  geom_smooth(method = "loess", se = F) + 
  geom_point(aes(color = Characteristic, size = Equity_in_Own_Home)) + 
  scale_x_continuous(limits = c(0, 350000),
                     breaks = c(0, 100000, 200000, 300000, 400000)) +
  scale_y_continuous(limits = c(0, NA),
                     breaks = c(0, 20000, 40000, 60000, 80000, 100000, 120000)) +
  scale_size_continuous(name = "Median Equity in House") +
  scale_color_discrete(name = "Highest Level of Education") + 
  labs(title = "Educational Attainment Retirement Accounts and Overall Networth",
       x = "Median Net Worth",
       y = "Median Total in Retirement Accounts",
       caption = "Source: Census Bureau") +
  theme(
    plot.title = element_text(face = "bold", size = 10),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold"),
    axis.line = element_line(color = "black"),
    panel.grid.minor = element_blank())
```

And for a third plot:
```{r plot3, fig.width= 12, fig.height=8, warning = FALSE,  echo = FALSE}
ggplot(fed_hh_balance_2000, mapping = aes(x = date)) + 
  geom_point(mapping = aes(y = disposable.personal.income), color = "blue") + 
  scale_x_discrete(breaks = c("2000:Q1", "2002:Q1", "2004:Q1", "2006:Q1", 
                              "2008:Q1", "2010:Q1", "2012:Q1", 
                              "2014:Q1", "2016:Q1", "2018:Q1")) + 
  labs(title = "Household Disposible Personal Income",
       x = "Date: 2000-2019",
       y = "Disposible Personal Income (in Millions of Dollars)",
       caption = "Source: Federal Reserve") +
  theme(
    plot.title = element_text(face = "bold", size = 10),
    axis.text.x = element_text(face = "bold", angle = 45),
    axis.text.y = element_text(face = "bold"),
    axis.line = element_line(color = "black"),
    panel.grid.minor = element_blank())
```